---
title: "Feline Urinary Obstruction Data Analysis"
date: "February 21, 2025"
output:
  pdf_document
header-includes:
  - \pagenumbering{gobble}
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
#| label: Loading packages
#| warning: false
#| message: false
library(mosaic)  
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)
library(writexl)
library(plotly)
library(dplyr)
library(readr)
library(knitr)
library(kableExtra)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: Tidying data
dataset <- read_excel("dataset.xls")
# Extracting relevant variables
Difficulty_Data <- dataset |>
  select(Age, `Weight (lbs)`, `Previous UO`, Difficulty, `Size of u-cath`, `Duration of signs (days)`, `Repeat UO?`) |>
  rename(Weight = `Weight (lbs)`,
         Hx = `Previous UO`,
         C_Size = `Size of u-cath`,
         Sx_Days = `Duration of signs (days)`,
         Repeat = `Repeat UO?`)

```

## I. Introduction

This report presents the results of an Exploratory Data Analysis (EDA)
conducted on a dataset related to feline urinary obstruction (UO). The
objective is to identify patterns and relationships in the data that can
help answer key research questions concerning factors influencing UO
procedure difficulty, recurrence, and post-treatment outcomes.

The dataset comprises 105 observations and 93 variables, including
numerical and categorical attributes related to patient demographics,
medical history, treatment procedures, and clinical outcomes. The
primary goal of this analysis is to extract insights that can inform
veterinary decision-making and improve patient care.

We are looking at a few variables to observe their relation to repeat
Urinary Obstructions in cats. These variables are: `Weight`, `Age`,
`History`, and `Difficulty`.

## II. Research Questions and Findings

#### **Were overweight patients more likely to have another urinary obstruction than their healthy-weight counterparts?**

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

Difficulty_Data |> summarise(favstats(Weight))

```

   

We found that the overweight cats were not more likely to have a
repeat UO. We separated the data into two weight classes under the
variable name `weight_class`. The 3rd quartile of patient's weights in
the dataset was 15.4 lbs, so we grouped cats who were over that
threshold as `Heavy` and those under the threshold as `Not Heavy`. The
proportion of cats who had a repeat UO and cats who didn't in both
groups was roughly the same, pointing towards no correlation.

```{r, echo=FALSE, message=FALSE, fig.width=6, fig.height=4}
Weight_Data <- Difficulty_Data |>
  mutate(weight_class = ifelse(Weight >= 15.4, "Heavy", "Not Heavy")) |>
  filter(Repeat != "NA")

counts <- Weight_Data |> 
  group_by(weight_class, Repeat) |>
  summarise(count = n())

counts <- setNames(counts, c("Weight Class", "Repeat UO", "Number of Cases"))
kable(counts, format ="latex") |>
        kable_classic(full_width = FALSE, latex_options = c("hold_position"))
  
```

#### **Were patients with histories of urinary obstruction more likely to have a difficult procedure?**

    We found no connection between cats with previous instances of
urinary obstruction and having a greater chance of having a difficult
procedure. The number of cats with a difficult procedure and those
with an easy one were nearly the same. Among the two groups the
proportion of cats with a history of UO was very similar. 12.6% of
the cats had a history, and around 46% of those cats had a difficult
procedure.

```{r, echo=FALSE, message=FALSE, results='hide', fig.width=6, fig.height=4}
Difficulty_Hx_Data <- Difficulty_Data |>
  filter(Hx != "Unknown")
  # Two cats had unknown histories of UO

Repeat_Hx_Stats <- Difficulty_Hx_Data |>
  group_by(Hx, Difficulty) |>
  summarise(count =n())

Repeat_Hx_Stats

Difficulty_Hx_Contingency <- addmargins(table(Difficulty_Hx_Data$Difficulty, Difficulty_Hx_Data$Hx))

Difficulty_Hx_Contingency_Props <- addmargins(prop.table(table(Difficulty_Hx_Data$Difficulty, Difficulty_Hx_Data$Hx)))

Difficulty_Hx_Contingency_Props

mosaicplot(data = Difficulty_Hx_Data, Difficulty ~ Hx, 
           color = c("#ADD8E6", "#AD001D"),
           xlab = "Difficulty",
           ylab = "Previous UO",
           main = "Procedure difficulty by previous urinary obstruction")

```

#### **Were patients that experienced difficult procedures more likely to have another urinary obstruction?**
\
Patients who experienced difficult procedures were not more likely to
have another urinary obstruction. The proportion of cats who had repeat
urinary obstruction as opposed to cats who did not have one was
extremely similar in both difficulty groups.

```{r, echo=FALSE, message=FALSE, results='hide', fig.width=6, fig.height=4}

Difficulty_Repeat_Data <- Difficulty_Data |>
  filter(Repeat != "NA") |>
  filter(Difficulty != "NA")

Repeat_Age_Stats <- Difficulty_Repeat_Data |>
  group_by(Repeat, Difficulty) |>
  summarise(favstats(Age)) |>
  select(Repeat, Difficulty, median, mean, n)

Repeat_Age_Stats

Difficulty_Repeat_Data |>
  group_by(Repeat) |>
  summarise(fav_stats(Age))
```

```{r, echo=FALSE, message=FALSE, fig.width=6, fig.height=4}
Repeat_Age_Stats <- setNames(Repeat_Age_Stats, c("Repeat", "Difficulty", "Median Age", "Mean Age", "Number of Cases"))
kable(Repeat_Age_Stats, format ="latex") |>
        kable_classic(full_width = FALSE, latex_options = c("hold_position"))

```

However, we did observe a relationship between a cat's age and
repeat urinary obstruction regardless of the procedure difficulty. Both
the median and mean age of cats who didn't have a repeat UO were lower
than cats who did have a repeat, regardless of the difficulty group they
were in. Other analysis methods (like a simple logistic regression)
reinforce this relationship. In a simple model we created to predict the probability of a cat having a repeat OU, where age was the only variable taken into account, age had a significant effect on the odds of
having a repeat OU. For every additional year a cat lives, its odds of
having a repeat urinary obstruction increase by around 29%.

```{r, echo=FALSE, message=FALSE, results='hide', fig.width=6, fig.height=4}
Difficulty_Data$Repeat_Binary <- ifelse(Difficulty_Data$Repeat == "Yes", 1, 0)

model <- glm(Repeat_Binary ~ Age, data = Difficulty_Data, family = binomial)
summary(model)

#odds ratio
exp(coef(model))

ggplot(data = Difficulty_Repeat_Data, aes(x = Difficulty, y = Age, fill = Repeat)) + 
  geom_boxplot() + 
  facet_wrap( ~ Repeat, ncol = 4) + 
  scale_fill_manual(values = c("#ADD8E6", "#AD001D")) +
  labs(title = "Age v. procedure difficulty by repeat UO") + 
  theme_minimal()

# FINDINGS:
#           1. Most cats (~87% of those for whom there was repeat data) did not get another UO after the procedure.
#           2. The median age of cats who had a subsequent UO was 8.4 years compared to 4.6 among cats who did not get another.

# It appears that there may be a relationship between age and repeat UO regardless of procedure difficulty, suggesting that younger cats fare better.
```
